---
title: "Impact of Major Storms - Fatalities, Injuries, and Economic"
author: "Bill Holst"
date: "January 12, 2016"
output:
  html_document:
    keep_md: true
---
##Synopsis of Storm Data Analysis

This analysis examines publicly available storm data, collected by the U.S. National Oceanic and Atmospheric Adminstration (NOAA) in their storm database. This data was collected from 1950 to 2011. 

The database, as described on the NOAA website, has interesting points about the data:

* Only tornado, thunderstorm, wind and hail events were recorded prior to 1996
* From 1996, there were 48 different event types recorded.

These facts point us to the fact that there is some data cleanup needed in the event classification to get meaninful results. Because there is incomplete data prior to 1996, this analysis only covers from 1996 to 2011. 
  
The purpose of this analysis is to reveal the major types of storm events and their impact on both the human population (deaths and injuries), and the economic impact (property and crop damage.)  

The results of the analysis are shown in the Analysis and Results section.

##Data Processing

First, we need to establish our work environment and download the file if we haven't done so already. The downloaded csv file is called 2F2FStormData.csv.bz2.

```{r cache=TRUE}

setwd("~/GitHub/Storm-Data-Analysis")

UrlName = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
csvfile = "2F2FStormData.csv.bz2"

# don't download file if we have already
if (!file.exists(csvfile)) {
    fileUrl = UrlName
    download.file(fileUrl,destfile=csvfile)
    
    dateDownloaded = date()
    dateDownloaded
}
# Read the data and take a look

    StormData = read.csv (csvfile)
    dim(StormData)
    str(StormData)

```

We only need a summary of this data, so we eliminate all but the 9 fields of interest for the analysis. These are the fields related to injuries, fatalities, and economic impact.

The NOA website at http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype informs us that prior to 1996, they only recorded tornado, thunderstorm, wind and hail events. To ensure consistent detail, we will use a subset of the data which includes only events from 1996 through 2011.


```{r}
StormData <- StormData[, c("BGN_DATE", "EVTYPE", "FATALITIES","INJURIES",                           "PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
dim(StormData)
summary(StormData$EVTYPE)
UniqueEvents = unique(StormData$EVTYPE)
noEvents = length(UniqueEvents)
```

There are `r noEvents` distinct values for event type. This list must be consolidated into fewer groups.

```{r}
# select only the fatalities and injuries
StormSubset = StormData[StormData$FATALITIES>0 | 
    StormData$INJURIES > 0 | StormData$PROPDMG > 0 |
    StormData$CROPDMG > 0, ]
StormSubset$Dt = as.POSIXlt(StormSubset$BGN_DATE,format="%m/%d/%Y %H:%M:%S")
# select the data since 1996
Since1996=StormSubset[format(StormSubset$Dt,"%Y") >= "1996",]
```

Apply a little cleanup to the EVTYPE field.

```{r}

trim <- function (x) gsub("^\\s+|\\s+$", "", x)
Since1996$EVTYPE = trim(Since1996$EVTYPE)
# convert all to upper
Since1996$EVTYPE = toupper(Since1996$EVTYPE)

```
From the above summary, the event type field is what we need to break the storm data into meaninful groups. However, the data is very dirty, e.g. TSTM WIND is most likely the same as THUNDERSTORM WIND. As a matter of fact, there are  There are 985 unique values - we should try to analyze around 8 to 10 major storm groupings, consolidating some of the obvious duplicates. 

Fortunately, a researcher, Coursea colleague, and R blogger named Mauricio Linhares has developed a substitiution table to eliminate the various misspellings and duplicate entries in the EVNTYPE field. His information about the process is at http://rpubs.com/jlutzwpi/56060. In the spirit of reproducible research, we choose to use his well-crafted substitution table to consolidate our events into a very workable number. 

The translation file that normalizes the event data is called replacements.csv, which we merge with the event data.

```{r}
url2 = 
    "https://raw.githubusercontent.com/mauricio/reproductible-research-assignment-2/master/replacements.csv"
if (!file.exists("replacements.csv")) {
    download.file (url2,"replacements.csv")
}
repl = read.csv("replacements.csv",stringsAsFactors = FALSE)
names(repl) = c("x","EVTYPE","Event")

# Merge these over the EVTYPE field  

Since1996Merged = merge(Since1996,repl,by="EVTYPE")
names(Since1996Merged) =  c(names(Since1996Merged[1:10]),"Event")
dim(Since1996Merged)

# group the data by total fatalities and injuries

FatalandInjuries =aggregate(
    x=Since1996Merged[,c("FATALITIES","INJURIES")],
    by=list(Since1996Merged$Event),FUN="sum")

# create the ordered fatality and injury data

SortedFatal=
    FatalandInjuries [order(-FatalandInjuries$FATALITIES),][1:10,]
SortedInjuries = 
    FatalandInjuries [order(-FatalandInjuries$INJURIES),][1:10,]
```
We also need to compute the cost of the events. First, some data cleanup must occur. The in both cases, dollar cost is codified by fields that indicate the magnitude of the cost: K-1000, M-millions and B-billions. A little math then calculates the cost of each.

```{r}
Since1996Merged$PropFactor = 1
Since1996Merged$PropFactor [Since1996Merged$PROPDMGEXP=="K"] = 1000
Since1996Merged$PropFactor [Since1996Merged$PROPDMGEXP=="M"] = 1000000
Since1996Merged$PropFactor [Since1996Merged$PROPDMGEXP=="B"] = 1000000000
Since1996Merged$CropFactor = 1
Since1996Merged$CropFactor [Since1996Merged$CROPDMGEXP=="K"] = 1000
Since1996Merged$CropFactor [Since1996Merged$CROPDMGEXP=="M"] = 1000000
Since1996Merged$CropFactor [Since1996Merged$CROPDMGEXP=="B"] = 1000000000
Since1996Merged$PropertyExpense = Since1996Merged$PROPDMG * Since1996Merged$PropFactor
Since1996Merged$CropExpense = Since1996Merged$CROPDMG * Since1996Merged$CropFactor
Since1996Merged$TotalExpense = 
    Since1996Merged$PropertyExpense + Since1996Merged$CropExpense

EventCosts = aggregate(
    x=Since1996Merged[
        ,c("PropertyExpense","CropExpense","TotalExpense")],
         by=list(Since1996Merged$Event),FUN="sum")

# pick the top 10 most expensive event catagories

SortedEventCosts=EventCosts[order(-EventCosts$TotalExpense),][1:10,]

```
## Results

The top 10 worst weather events from 1996 to 2011 are shown below, both in terms of fatalities and injuries:

```{r fig.width=6, fig.height=7}
library(ggplot2)
library(gridExtra)

p1 = ggplot(data=SortedFatal,
        aes(x=reorder(Group.1, -FATALITIES),
        y=FATALITIES,fill=FATALITIES)) +  
        geom_bar(stat="identity", width = 0.5) +
        xlab("Event") +
        ylab("Total Number of Fatalities") +
        theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1)) +
        ggtitle("Top 10 Weather Event Fatalities 1996-2011")
        
p2 = ggplot(data=SortedInjuries, height = 5,
        aes(x=reorder(Group.1, -INJURIES),
        y=INJURIES,fill=INJURIES)) +  
        geom_bar(stat="identity", width = 0.5) +
        xlab("Event") +
        ylab("Total Number of Injuries") +
        theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1)) +
        ggtitle("Top 10 Weather Event Injuries 1996-2011") 

grid.arrange( p1, p2, nrow=2)
```        

Now, we need to analyze the cost of the major events. First, we look at the sorted expenses.


```{r}
SortedEventCosts

```
The costs in this analysis have not been normalized for inflation. That is beyond the scope of the exercise. 

In general, the property damage exceeds crop damage by considerable amount. We show both in a stacked bar chart. The following figure shows total storm-related expense by most expensive event:

```{r fig.width=6, fig.height=7}
library (reshape2)
SortedEventCosts=EventCosts[order(-EventCosts$TotalExpense),][1:10,]
Melted=melt(SortedEventCosts, id.var="Group.1")
Melted = Melted[Melted$variable != "TotalExpense",]
names(Melted) = c("Event","Type","Expense")
ggplot(Melted,aes(x=reorder(Event,-Expense),
    y=Expense/1000000000,fill=Type)) +
    geom_bar(stat="identity") +
    ylab("Cost (in $Billions)") +
    xlab("Event") +
    theme (legend.position="none",
    axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle("Top 10 Storm-Related Costs 1996-2011") 
```

### Summary and Conclusions

Reviewing the graphs, we find that tornados play a major factor in terms of injuries and death, ranking first in injuries and second in deaths. Heat-related situations result in the most deaths, and rank second in injuries. Overall cost of events find that floods are the largest contributor to economic cost of storm-related events. Both floods and flash floods are among the top six events.  Drought is the only event that contributes primarily to crop damage. 

One point of interest is the relationship between hurricanes and floods. Hurricane Katrina was the most expensive natural disaster in U.S. history, with costs exceeding $126 billion, with at least half of the cost the related flooding following the levy breaks 
(see http://useconomy.about.com/od/grossdomesticproduct/f/katrina_damage.htm.)  It is not clear from this analysis whether those costs are included only in the hurricane event or split over the flood events too.

An interesting follow up to this analysis would be to find out how hurricane and flood events were reported, to see if hurricanes might be a root cause of flood reporting.

## References

NOAA Website - http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype

Event Cross Reference - http://rpubs.com/jlutzwpi/56060

Data Location - "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

Coursera Project Information - https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2

Katrina Costs: http://useconomy.about.com/od/grossdomesticproduct/f/katrina_damage.htm